---
title: "ggplot2 Tutorial"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article includes a brief explanation of how to use `ggplot2`, and how to integrate the **seizer** functions with it. This is far from an in-depth article, but it should be enough for you to be able to generate some basic plots using `ggplot2`. See the [Further reading] section if you want to dive deeper into the world of `ggplot2`.

We'll begin, always, by loading our packages:

```{r setup, warning = FALSE}
library(ggplot2)
library(seizer)
```

## The grammar of graphics

The key to using `ggplot2` is understanding its syntax, which is based on Hadley Wickham's layered grammar of graphics framework. If you want to read more about it, there are links in the [Further reading] section. For our purposes though, we'll just focus on the components that make up a `ggplot` object, and how we set it up. Here is a representation of the structure of a `ggplot` object (adapted from the `ggplot2` documentation):

```{r grammar, echo=FALSE}
n <- 7
x <- outer(c(-2, 0, 2, 0), rep(1, n))
y <- outer(c(0, 1, 0, -1), seq(0, 2.309, length.out = n), FUN = `+`)

df <- data.frame(
    x = as.vector(x),
    y = as.vector(y),
    group = as.vector(col(x))
)

ggplot(df, aes(x, y, group = group, fill = factor(group))) +
    geom_polygon(alpha = 0.9) +
    coord_equal() +
    scale_y_continuous(
        breaks = seq(0, 2.309, length.out = n),
        labels = c("Data", "Mapping", "Layers", "Scales",
                   "Facets", "Coordinates", "Theme")
    ) +
    scale_fill_cesar_d(
        guide = "none"
    ) +
    theme_cesar() +
    theme(axis.text.y = element_text(face = "bold", hjust = 1, size = 14),
          axis.text.x = element_blank(),
          axis.title = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.major.y = element_blank())
```

As you can see, the key word in the layered grammar of graphics behind `ggplot2` is *layered*. Each component is passed separately to a function, and you add them one to another to create a complete plot. We will explain how by recreating the following figure stage by stage:

```{r ggplot cesar scale, message = FALSE, warning = FALSE, echo = FALSE}
iris %>% ggplot(aes(x = Sepal.Length, y = Petal.Length, colour = Sepal.Width, fill = Species)) + geom_point() + geom_smooth(colour = ancient_lavastone, method = "lm") + facet_wrap(~Species, scales = "free") + labs(title = "This is a title", subtitle = "This is a subtitle") +
  theme_cesar() + scale_colour_cesar_c(palette = "galliano_c") + scale_fill_cesar_d()
```

### 1. Data

The basis of any plot is the data underlying it. Our data will consist of a dataframe, and we will want to visualise at least one dimension, or column, in the data. For example, we can view the `iris` dataset which comes together with `ggplot2` and has data of sepal and petal dimensions for different species of iris flowers.

```{r iris}
head(iris)
```

Irises, by the way, are awesome and you should take this opportunity to google some photos of irises. I'm a particular fan of *Iris petrana*.

Anyway, moving on!

We can generate a `ggplot` object by passing data to the `ggplot2::ggplot()` function. This can be done either by using the `data` argument in the `ggplot` function:

```{r data 1}
p <- ggplot(data = iris)

p
```

or by piping, if you're comfortable using that functionality:

```{r data 2}
p <- iris %>% ggplot()

p
```

Note that in both cases, we end up with an empty plot. This is because while we have provided `ggplot2` with the data, we have not provided any other information on how to construct our plot. But we have created a `ggplot` object, which we can now add more and more to using other functions.

### 2. Mapping

The next step is to decide which dimension, or dimensions, we want to plot. In other words, we define the axes of our plot, and whether any dimensions are encoded to size, shape, colour, etc. This is done using the `ggplot2::aes()` function, which defines a list of aesthetic mappings, and is passed on to the `ggplot` object using the `mapping` argument. This is how we translate the data to the graphics system. So, for example, if we want to plot petal length against sepal length, we pass these as our axes to the `ggplot` object:

```{r aes}
p <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length))

p
```

Our plot is coming along - we now have axes, and we see that the ranges on the axes correspond to the range of values in each column in the dataframe. But the plot is still empty.

### 3. Layers

Layers display the mapped data from the previous step in some kind of graphical representation. To quote from the `ggplot` documentation:

Every layer consists of three important parts:

1. The **geometry** that determines how data are displayed, such as points, lines, or rectangles.
2. The **statistical** transformation that may compute new variables from the data and affect what of the data is displayed.
3. The **position** adjustment that primarily determines where a piece of data is being displayed.

Layers are constructed using either `geom_*()` or `stat_*()` functions. We will focus here on geoms, which are the geometric objects with which we want the data visualised. These can be points, bars, lines, boxes, etc. There are many different geoms to choose from - but some will not work with our data. For example, this works fine:

```{r geom 1}
p + geom_point()
```

but this throws up an error:

```{r geom 2, error = TRUE, fig.show = "hide"}
p + geom_bar()
```

You'll notice the error mentions something called `ggplot2::stat_count()`. This is the statistical transformation - which determines whether or not we want to show some kind of statistical measure of our underlying data, for example a count. But it could also be the mean, spread, confidence intervals, etc. All geoms have underlying default `stat` values, which define the statistical transformation to use on the data in this layer. But some combinations just don't make sense. So, our first attempt worked because `ggplot2::geom_point()` by default uses `stat = "identity"` - which means the data are not transformed, and each datum is shown as a point. But, `ggplot2::geom_bar()` by default uses `stat = "count"`, which doesn't work with two dimensions. This is a statistical transformation that counts the number of unique values in a column, so it really only works when you have a single dimension, for example:

```{r geom 3}
ggplot(iris, aes(x = Sepal.Length)) + geom_bar()
```

However, some geoms with `stat` transformations other than `identity` can work very neatly with two dimensions. For example, if we want to add a trend line to our scatterplot, we can use `ggplot2::geom_smooth()`, which shows the central tendency and confidence intervals of the y variable along the range of the x axis. Sounds complicated, but really - it's just the output of a model where y is predicted by x:

```{r smooth 1}
p <- p + geom_point()

p + geom_smooth()
```

By default, `ggplot2::geom_smooth()` fits a non-linear model (either `loess` or `gam` depending on the sample size of the data), but we can constrain it to show a linear regression - which means we have now created a scatterplot with a trend line from a linear model which we could construct using the code `lm(Petal.Length ~ Sepal.Length, data = iris)`:

```{r smooth 2}
p <- p + geom_smooth(method = "lm")

p
```

An important thing to note here - each `geom_*()` function defines a **new layer**. This means we can potentially use different aesthetic mappings, or even different data, for a new geom. To explore what that means, we'll start by assigning an aesthetic mapping to colour:

```{r scale colour 1}
p <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Species)) + geom_point() + geom_smooth(method = "lm")

p
```

We've now mapped the colour aesthetic onto the column Species, so that each unique value of Species gets its own colour. This applies to all geoms, because we passed the colour aesthetic to the `ggplot2::ggplot()` function. Hence, we get three different regression lines, because by mapping colour to a discrete variable, we have also implicitly passed a grouping variable - *i.e.*, the statistical transformation underlying `ggplot2::geom_smooth()` is applied separately to each group.

However, we could also map the aesthetic variable to a single layer, by passing it to only one `geom_*()`:

```{r scale colour 2}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) + geom_point(aes(colour = Species)) + geom_smooth(method = "lm")
```

One important thing to note here is that unlike in the previous plot, the regression line is shared. This is because the colour aesthetic mapping is not passed to `ggplot2::geom_smooth()`, and so it doesn't include a grouping variable in the statistical transformation. You can pass it separately to `ggplot2::geom_smooth()` using the `group` aesthetic mapping, so that you still get separate regression lines but without individual colours:

```{r scale colour 3}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) + geom_point(aes(colour = Species)) + geom_smooth(aes(group = Species), method = "lm")
```

This is just an example of the strength of the layered structure of `ggplot2`.

For more expert `ggplot2` users, you can use `geom_*()` in unconventional ways for more complex data vis by changing the `stat` value, or by passing a `stat_*()` function and adjusting its `geom` argument. This is beyond the scope of this tutorial, but feel free to experiment if you're interested.

Finally, there is the matter of the position. Just as with statistical transformations, each `geom_*()` function comes with its own default `position` argument. We won't get into it here, but know that this can be a very useful tool for some visualisations by nudging components to the sides or overlapping them.

At this point, we have everything we need to create a simple plot. But there's a lot more we can do by adjusting the next four layers in our grammar of graphics cake.

### 4. Scales

The Scales component refers to scaling values in the plot, or using specific scales to represent multiple values or a range. One basic usage is to make changes to our axes. Scale functions are patterned as `scale_{aesthetic}_{type}()`, where `{aesthetic}` is one of the pairings made in the mapping part of a plot. So if we want restrict the range of values by setting new limits on the axis:

```{r scale 1}
p + scale_y_continuous(limits = c(4, 7))
```

**Important:** *note the warnings! Data have been removed, which changes the statistical transformation and also drops an entire species from the plot! We'll get into what this means when we get to discussing coordinate systems.*

We could also use the `scale_*()` function to transform the axis. For example, if we want to log an axis:

```{r scale 2}
p + scale_y_continuous(transform = "log")
```

Another good use - we can rename a dimension using the `scale_*()` function. Like this:

```{r scale 3}
p + scale_x_continuous(name = "New name")
```

Note the distinction between discrete and continuous types, the third element in  `scale_{aesthetic}_{type}()`. If a discrete variable is mapped to one your axes you can only transform it using `scale_{aesthetic}_discrete()`. Attempting to use `scale_{aesthetic}_continuous()` like in the previous example will generate an error:

```{r scale discrete, error = TRUE, fig.show= 'hide'}
ggplot(iris, aes(x = Species, y = Petal.Length)) + geom_point() + scale_x_continuous()
```

Finally, we can use `scale_*()` functions to assign new representations to aesthetic mappings other than the axes. For instance, we can use `scale_*()` functions to change the size, shape or colour values used to represent specific values of the underlying data. For a colour scale, that means we can assign different colours to the mapped groups:

```{r scale colour 4}
p + scale_colour_manual(values = c("red","blue","purple"))
```

Or, we can use our Cesar Australia colour palettes by using one of the custom functions included with **seizer**:

```{r scale colour 5}
p + scale_colour_cesar_d()
```

Note we use `scale_colour_cesar_d()` for a discrete colour palette, because colour is mapped to a character variable. If we map colour to a continuous variable, such as sepal width, we need to use the appropriate function, `scale_colour_cesar_c()`.

```{r scale colour 6}
p <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Sepal.Width)) + geom_point() + geom_smooth(method = "lm")

p + scale_colour_cesar_c(palette = "galliano_c")
```

Mind the warning that popped up - this refers to the `colour` aesthetic being dropped from `ggplot2::geom_smooth()` because it is meaningless to assign a continuous colour mapping to this geom.

One important thing to note - you cannot include more than one colour scale*. However, there is another aesthetic mapping which assigns colours - fill. The difference between them is that `colour` refers to outlines, and `fill` to, well, fills. Some geoms do not have `fill` values - for example lines, or points, which are all one dimnensional. But some do - for example, polygons like the confidence intervals around our `ggplot2::geom_smooth()`. So, we can assign Species to `fill` to have colour-coded species while also assigning colours to different values of our third variable, sepal width.

<font size="1">* unless you use the [ggnewscale](https://eliocamp.github.io/ggnewscale/) package</font>

```{r scale colour 7}
p <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Sepal.Width, fill = Species)) + geom_point() + geom_smooth(method = "lm")

p + scale_colour_cesar_c(palette = "galliano_c") + scale_fill_cesar_d()
```

To truly see the difference between `colour` and `fill` it is best to look at a `geom` that has both. Compare this:

```{r scale colour 8}
ggplot(iris, aes(x = Species, y = Petal.Length, colour = Species)) + geom_boxplot() 
```

To this:

```{r scale colour 9}
ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) + geom_boxplot() 
```

One final note about colour - we can also assign colours to geoms without mapping them to variables. For example, if we just want to change the colour of the lines so they're not blue, we can do this by assigning a value to the `colour` argument of `ggplot2::geom_smooth()` *outside* of the `ggplot2::aes()` function (in this example we use `ancient_lavastone`, which is a colour included with the **seizer** package):

```{r scale colour 10}
p <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Sepal.Width, fill = Species)) + geom_point() + geom_smooth(colour = ancient_lavastone, method = "lm") + scale_colour_cesar_c(palette = "galliano_c") + scale_fill_cesar_d()

p
```

### 5. Facets

Ok, we're starting to get close to the final product. We will now discuss facets, which are a way of dividing our plot to subplots by using subsets of the data, defined by variables of a given column. `ggplot2::facet_*()` functions can accept a mapping of variables as a formula of the form `y ~ x` where `y` represents the variable coded to rows (each value is displayed in a separate row), and `x` the variable coded to columns (each value is displayed in a separate column). We can omit either `y` or `x` if we just want subplots across columns or rows, respectively. For example, if we want to divide our plot into a subplot for each species next to one another (separate columns):

```{r facets 1}
p + facet_wrap(~ Species)
```

Note that all three facets have shared ranges on their axes. This can sometimes be useful, if you want to easily compare one group to another. In other cases, it may be informative - for instance, if the range of values in one group is so large that the variation in another is masked by it. We can sort of see something like that happen with the species *setosa*, where most of the subplot is empty space. We can get around this using the `scales` argument in the `facet_*()` function. We can free up one or both of the axes by passing `scales = "free_x"`, `scales = "free_y"` or `scales = "free"`, which essentially changes the scale parameters of individual facets based on the data range for the group in the data subset:

```{r facets 2}
p <- p + facet_wrap(~ Species, scales = "free_x")

p
```

### 6. Coordinates

The final component is the coordinate system. These define how your position aesthetics (x and y) are interpreted and displayed. You basically have two options* here - Cartesian (the default), and polar. Basically, Cartesian coordinate system is where your x and y axes are perpendicular. In a polar system, each point is defined by a radial coordinate (distance from the centre) and an angular coordinate (or azimuth). Sometimes data vis will require polar coordinates (*e.g.*, pie chart), but you will likely never have to worry about it. But if you want to see what happens when you use a polar coordinate system in `ggplot2`:

<font size="1">* unless you're creating maps, in which case [coord_sf](https://ggplot2.tidyverse.org/reference/ggsf.html) gets into the picture</font>

```{r polar}
p + coord_polar()
```

Looks cool eh? Pretty nonsensical from a data vis perspective, but we can see what happened here - Petal.Length has become the axis along the radius (radial coordinate) and Sepal.Length has become the azimuth (angular coordinate).

Anyway, the main use you will get from playing with coordinates is if you want to zoom in on particular parts of the data. Remember when we changed the limits in our axis scale? Here it is again:

```{r cartesian 1}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Species)) + geom_point() + geom_smooth(method = "lm") + scale_y_continuous(limits = c(4, 7))
```

Here's what happens if you do something similar using the function `ggplot2::coord_cartesian()`:

```{r cartesian 2}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Species)) + geom_point() + geom_smooth(method = "lm") + coord_cartesian(ylim = c(4, 7))
```

See the difference in the green line? This goes back to those warnings we mentioned when changing the range of the axis using a `scale_*()` function. `scale` is a transformation. You're telling `ggplot2` to only consider data within the range defined by the `scale_*()` function - so the regression function underlying `ggplot2::geom_smooth()` only uses data within that range to calculate the trend line. Coordinates, however, only change how the data are displayed, not how anything is calculated. So this gives us a real zoom in, without actually changing any of the data underlying the plot.

### 7. Theme

This is another important way to play around with `ggplot2` objects. `ggplot2::theme()` controls all of the static elements of a plot - basically anything to do with how it looks that isn't controlled by the underlying data. This can be things like font size of labels, width of grid lines, background colour, etc. For example, if you want to get rid of a legend it's as simple as running:

```{r legend}
p + theme(legend.position = "none")
```

This is actually an important component of data vis, since this can have a large impact on how visually appealing and neat your plot looks. Otherwise good data vis can fail if the plot comes out looking too cluttered, and the static elements can have a lot to do with this. From a company or team perspective, this can also be where you standardise your plots to a distinct visual style that reinforces your brand identity. Unfortunately, while powerful and important, `ggplot2::theme()` is also very complicated. There are **a lot** of parts you can customise. So, to save you the trouble, **seizer** comes with its own, inbuilt `theme_cesar()` function which makes your figure look nice and tidy and in line with the Cesar Australia style guide!

```{r theme}
p <- p + theme_cesar()

p
```

We've almost recreated our figure. The one thing we're missing is labels - and while you can change those using `scale_*()` or `ggplot2::theme()` functions, you can also use the handy `ggplot2::labs()` function. Note that you can use this to add titles and susbtitles, but also to change the names of existing axes and/or aesthetic mappings.

```{r labs}
p <- p + labs(title = "This is a title", subtitle = "This is a subtitle")

p
```

So there we go! We have now successfully recreated the figure, using all of our distinct stages, apart from Coordinates which we leave at default values:

```{r full, fig.show = "hide", message = FALSE, warning = FALSE}
ggplot(iris, # 1. Data
       aes(x = Sepal.Length, y = Petal.Length, colour = Sepal.Width, fill = Species)) + # 2. Mapping
  geom_point() + geom_smooth(colour = ancient_lavastone, method = "lm") + # 3. Layers
  scale_colour_cesar_c(palette = "galliano_c") + scale_fill_cesar_d() + # 4. Scales
  facet_wrap(~ Species, scales = "free_x") + # 5. Facets
  theme_cesar() + # 7. Theme
  labs(title = "This is a title", subtitle = "This is a subtitle") 
```


## Further reading

Well done! You are now well versed in the basics of `ggplot2` and should be able to build your own plots from scratch. However, if you feel like you need more help, or just want to delve deeper, you can explore the following resources:

[ggplot2: Get started](https://ggplot2.tidyverse.org/articles/ggplot2.html)

[The Complete ggplot2 Tutorial](https://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html)

[A layered grammar of graphics](https://vita.had.co.nz/papers/layered-grammar.html)

[A Comprehensive Guide to the Grammar of Graphics for Effective Visualization of Multi-dimensional Data](https://towardsdatascience.com/a-comprehensive-guide-to-the-grammar-of-graphics-for-effective-visualization-of-multi-dimensional-1f92b4ed4149)
